EGGs are adhesive!\\ An algebraic account of equality saturation

### 2025-09-08 jlamp/jlamp.tex
sez. 4

\newcommand{\pbc}{\mathsf{Pb}}
\newcommand{\pbe}{\mathsf{ePb}}
\usepackage{amsthm,amsmath,amssymb,mathrsfs, dsfont}
%\usepackage{tikz-cd}
\usepackage{hyperref, cleveref} %simboli matematici
\usepackage[all, cmtip]{xy}
\usepackage{thm-restate}

@@

\def\R{\mathsf{R}}
\def\B{\textbf {\textup{B}}}
\def\C{\textbf {\textup{C}}}
\def\D{\textbf {\textup{D}}}
\def\X{\textbf {\textup{X}}}
\def\Y{\textbf {\textup{Y}}}
\def\E{\textbf {\textup{E}}}
\def\T{\textbf {\textup{1}}}
\def\A{\textbf {\textup{A}}}
\def\M{\mathcal{M}}
%categorie varie
\newcommand{\catname}[1]{\textbf{\textup{#1}}}
\newcommand{\lab}{\catname{LHyp}}
\newcommand{\lmo}{\catname{LMH}}
\newcommand{\hyp}{\catname{Hyp}}
\newcommand{\phyp}{\catname{pHyp}}
\newcommand{\pshyp}{\catname{pSep}}
\newcommand{\shyp}{\catname{Sep}}
\newcommand{\hyps}{\catname{Hyp}_{\Sigma}}
\newcommand{\EqHyp}{\catname{EqHyp}} %equivalence hypergraphs
\newcommand{\EqHyps}{\catname{EqHyp}_{\Sigma}}
\newcommand{\EqTG}{\catname{EqTG}}
\newcommand{\EqTGs}{\catname{EqTG}_{\Sigma}}
\newcommand{\GEqTGs}{\catname{GEqTG}_{\Sigma}}
\newcommand{\GEqATGs}{\catname{GEqATG}_{\Sigma}}
\newcommand{\gr}{\textbf{\textup{Graph}}}
\newcommand{\dgr}{\catname{SGraph}}
\newcommand{\dg}{\catname{DAG}}
\newcommand{\rt}{\mathsf{dcl_s}}
\newcommand{\rta}{\mathsf{dcl}}
\newcommand{\rtd}{\mathsf{dcl_{d}}}
\newcommand{\slice}[2]{(\catname{#1}\downarrow{#2})}
\newcommand{\tg}[0]{\catname{TG}_{\Sigma}}
\newcommand{\teg}[0]{\catname{TeGr}_{\Sigma}}
\newcommand{\sv}[0]{\mathsf{Sieves}}
\newcommand{\mono}[1]{\mathsf{Mon}(\catname{#1})}
\newcommand{\mo}[1]{{#1}_\mathsf{Mon}}
\newcommand{\pro}{\mathsf{prod}}
\newcommand{\spro}{\mathsf{ps}}
\newcommand{\prol}{\mathsf{lprod}}
\newcommand{\pred}[1]{{\downarrow}#1}
\newcommand{\colim}[0]{\mathrm{colim}}
\newcommand{\cod}{\mathsf{cod}}
\renewcommand{\sp}{\mathsf{sp}}
\newcommand{\is}[1]{{#1}_{is}}
\renewcommand{\sup}{\mathsf{sup}}
\newcommand{\cow}[1]{\mathsf{cwd}({#1})}
\renewcommand{\inf}{\mathsf{inf}}
\newcommand{\dom}{\mathsf{dom}}
\newcommand{\dwnarrow}{\downarrow \hspace{-2pt}}
\newcommand{\Dwnarrow}{\Downarrow \hspace{-2pt}}
\newcommand{\egg}{\mathsf{e}\text{-}\catname{EqHyp}}

@@

\noindent 
\begin{minipage}[l]{.85\linewidth}
\begin{defi} A \emph{kernel pair} for an arrow $f\colon A \to B$ is an object $K_f$ together with two arrows $\pi^1_f, \pi^2_f\colon K_f \rightrightarrows A$, denoted as $(K_f, \pi^1_f, \pi^2_f)$, such that the square aside is a pullback.
\end{defi}
\end{minipage} \hfill 
\begin{minipage}[r]{.18\linewidth}
	\xymatrix@R=15pt{K_f \ar[r]^{\pi^2_f} \ar[d]_{\pi^1_f}& A \ar[d]^f \\ A \ar[r]^{f}& B}
\end{minipage}

@@

\smallskip 
\noindent
\begin{minipage}[l]{.75\linewidth}
	\begin{defi}
		A \emph{(single typed) monoidal signature} $\Sigma$ is a triple $(O_\Sigma, \ari_\Sigma, \coar_\Sigma)$ given by a \emph{set of operations} $O_\Sigma$ together with \emph{arity} and \emph{coarity} functions $\ari_\Sigma, \coar_\Sigma\colon O_\Sigma \rightrightarrows \mathbb{N}$.

@@

\subsubsection{Two  subcategories of $\hyp$}\label{sub:sub}

We are now going to introduce two classes of hypergraphs, and the relative subcategories, which will provide the base for our definitions of EGGs  (cfr.~\Cref{sec:sepegg,sec:lmegg}).

@@

\begin{cor}	$\lmo$ is closed in $\hyp$ under pullbacks.
\end{cor}

@@

\paragraph{Separated hypergraphs}
We can now introduce the second  subcategory of $\hyp$ in which we are interested in and that is going to be used in \Cref{sec:sepegg} to give the underlying hypergraphs for EGGs with minimally identified tokens.

@@

@@

We define the \emph{separated graph associated to $\mathcal{G}$}  as $\sp(\mathcal{G}) = (E_{\mathcal{G}}, V_{\sp(\mathcal{G})}, s_{\sp(\mathcal{G})}, t_{\sp(\mathcal{G})})$.
\end{defi}
\begin{rem}\label{rem:ob}
 By construction $\sp(\mathcal{G})$ is an object of $\shyp$. To see this it is enough to apply \Cref{rem:elem2}
 \begin{itemize}
 	\item every letter of $s_{\mathcal{G}}(e)$ and $t_{\mathcal{G}}(e)$ lies in the image of a different coprojection $\iota_e\circ \iota^e_i$, so it cannot appear more than one time;
 	\item every letter of $s_{\mathcal{G}}(e)$ lies in the image of $\iota_{e}\circ \iota^e_i$  for $i\leq n_e$, while every letter of $t_{\mathcal{G}}(e)$ lies in the image of $\iota_{e}\circ \iota^e_j$  for $n_e < j$, guaranteeing disjointness;
 	\item if $e$ is different from $e'$, then all the letters of $s_{\mathcal{G}}(e)$ and $t_{\mathcal{G}}(e)$ are  in the image of $\iota_e$, while those of  $s_{\mathcal{G}}(e')$ and $t_{\mathcal{G}}(e')$ are in the image of $\iota_{e'}$ and so, again, we get the required disjointness. 
 \end{itemize}
\end{rem}

@@

\begin{rem}\label{rem:mprun}
	Let $(f,g)$ and $(h, k)$ be two morphisms $\mathcal{G}\rightrightarrows \mathcal{H}$ of $\phyp$ and suppose that $f=h$. We can use the absence of isolated nodes to show that $g=k$ too. Let $v\in V_{\mathcal{G}}$, we have two cases:
	\begin{itemize}
		\item $v$ occurs in $s_{\mathcal{G}}(e)$ for some $e\in E_{\mathcal{G}}$. Let $1\leq i\leq \lgh_{V_{\mathcal{G}}}(s_{\mathcal{G}}(e))$ be such that $v=s_{\mathcal{G}}(e)_i$, then we have
		\[g(v)=g(s_{\mathcal{G}}(e)_i)=(g^\star(s_{\mathcal{G}}(e)))_i=s_{\mathcal{H}}(f(e))_i=s_{\mathcal{H}}(h(e))_i=(k^\star(s_{\mathcal{G}}(e)))_i=k(s_{\mathcal{G}}(e)_i)=k(v)\]
		\item $v$ occurs in $t_{\mathcal{H}}(e)$ for some $e\in E_{\mathcal{H}}$. Let $1\leq i\leq \lgh_{V_{\mathcal{H}}}(t_{\mathcal{G}}(e))$ be such that $v=t_{\mathcal{G}}(e)_i$, then we have
		\[g(v)=g(t_{\mathcal{G}}(e)_i)=(g^\star(t_{\mathcal{G}}(e)))_i=t_{\mathcal{H}}(f(e))_i=t_{\mathcal{H}}(h(e))_i=(k^\star(t_{\mathcal{G}}(e)))_i=k(t_{\mathcal{G}}(e)_i)=k(v)\]
	\end{itemize}
	In both cases $\hat{g}(v)=k(v)$ and so $\hat{g}= k$ as wanted.
\end{rem}

\begin{restatable}{prop}{dx}\label{prop:dx}
	The inclusion functor $P\colon \pshyp\to \phyp$ has a right adjoint $\sp_P$.
\end{restatable}

@@

If we restrict to pruned separated hypergraph the previous issue disappeares.  The crucial observation is that the a pruned separated hypergraph is characterized by its set of hyperedges and the length of their sources and targets. See \Cref{proof:pruned} for a proof.

@@

@@

\begin{defi}
	Let $\mathcal{G}$ be an hypergraph and $(h,k)\colon \mathcal{H}\to \mathcal{G}$  a $\mathcal{G}$-hypergraph. We say that $(h,k)$ is a \emph{left-monogamous} (resp.~\emph{separated}) $\mathcal{G}$-hypergraph if $\mathcal{H}$ is left-monogamous (resp.~separated).

	We will denote by $\lmo_{\mathcal{G}}$ and $\shyp_{\mathcal{G}}$  the corresponding subcategories of $\hyp_{\mathcal{G}}$.
\end{defi}

We can easily realize $\lmo_{\mathcal{G}}$ and $\shyp_{\mathcal{G}}$ as comma categories.

\begin{prop}
Let $M\colon \lmo \to \hyp$ and $S\colon \shyp \to \hyp$ be the inclusion functors. Let, moreover, $\Delta_{\mathcal{G}}\colon 1\to \hyp$ be the constant functor in $\mathcal{G}$. Then
\begin{enumerate}
	\item $\lmo_{\mathcal{G}}$ is equivalent to $\comma{M}{\Delta_\mathcal{G}}$;
	\item $\shyp_{\mathcal{G}}$ is equivalent to $\comma{S}{\Delta_\mathcal{G}}$; 
\end{enumerate}
\end{prop}

@@

\begin{thm} Let $\mathcal{G}$ be an hypergraph, then %the following fact hold
%	\begin{enumerate}
%		\item an arrow $(f,g)$  if and only if is so in $\lmo$; \marginpar{ovvero?}
		%\item 
		$\lmo_{\mathcal{G}}$ is quasiadhesive.
%	\end{enumerate}
\end{thm}

@@

@@

\Cref{thm:slice-functors,prop:mono}, \Cref{lem:reg} and \Cref{prop:push} allow us to recover the following result, previously proved by direct computation in \cite[Thm.~4.2]{CorradiniG05} (see also \cite[Cor.~5.15]{CastelnovoGM24} for the details of the argument presented here).
\begin{cor}\label{cor:term}
	The category $\tg$ is quasiadhesive.
\end{cor}
\fi 
\section{Adding equivalences to hypergraphical structures}
\label{hypereq}
The use of hypergraphs equipped with an equivalence relation over their nodes has been argued as a convenient way to express concurrency in the DPO approach to rewriting~\cite{concur2006}.
This section presents the framework by means of adhesive categories, including also its version for term graphs, as a stepping stone towards an analogous characterisation for e-graphs.

@@

\begin{restatable}{lem}{pbs}\label{lemma:stab}
	In $\EqHyp$, $\pbc$-pushouts are stable.
\end{restatable}

\begin{exa}
It is noteworthy to show that pushouts along regular monos are not stable.

@@

		\indent Consider e.g. the cube aside: the vertices are just graphs without edges, and in the graphs themselves the equivalence classes 

@@

the top face fails to be a pushout

@@

                \xymatrix@C=10pt@R=1pt{

@@

\subsubsection{title}
\paragraph{title}
\paragraph{title}
%\input{Hypergraphs_with_equivalences}
%\input{Labeled_hyp_with_eq}

@@

\begin{defi}
	Let $\Sigma$ be an algebraic signature and $\mathcal{G}^{\Sigma}$ the hypergraph associated to it. A \emph{labelled hypergraph with equivalence} is a pair $(\mathcal{H}, l)$ where $\mathcal{H}$ is an object of $\EqHyp$ and $l$ a morphism $T(\mathcal{H})\to \mathcal{G}^\Sigma$ of $\hyp$. A \emph{morphism of labelled hypergraphs with equivalence} between $(\mathcal{H}, l)$ and $(\mathcal{H}', l')$ is an arrow $h\colon \mathcal{H}\to \mathcal{H}'$ such that $l= l'\circ T(h)$.

	We denote the resulting category by $\EqHyp_\Sigma$.
\end{defi}

Let $(\mathcal{H}, l)$ be an object of $\EqHyp_\Sigma$: since $T$ has a right adjoint $R$ by \Cref{prop:forghyp}, $l\colon T(\mathcal{H})\to \mathcal{G}^\Sigma$ corresponds to the arrow $(l, !_{Q_{\mathcal{H}}})\colon \mathcal{H}\to R(\mathcal{G}^\Sigma)$. It is immediate to see that this correspondence extends to an equivalence with the slice over $R(\mathcal{G}^\Sigma)$.

\begin{prop}\label{prop:slice}
$\EqHyp_\Sigma$ is equivalent to $\EqHyp/R(\mathcal{G}^\Sigma)$.
\end{prop}

Let $V_\Sigma\colon \EqHyp_\Sigma\to \EqHyp$ be the functor forgetting the labelling and $\pbc_\Sigma$ the class of morphisms in $\EqHyp_\Sigma$ whose image in $\EqHyp$ is in $\pbc$. From \Cref{prop:eqhyp_complete,cor:pbapp}, the second point of \Cref{thm:slice-functors,cor:eqade}, we can deduce the following.

>>>>

\newcommand{\pbc}{\mathsf{Pb}}
\newcommand{\pbe}{\mathsf{ePb}}
\newcommand{\pbl}{\mathsf{LPb}}
\newcommand{\pble}{\mathsf{eLPb}}
\newcommand{\pbs}{\mathsf{SPb}}
\newcommand{\pbse}{\mathsf{eSPb}}
\usepackage{amsthm,amsmath,amssymb,mathrsfs, dsfont}
%\usepackage{tikz-cd}
\usepackage{hyperref, cleveref} %simboli matematici
\usepackage[all, cmtip]{xy}
\usepackage{thm-restate}

@@

\def\R{\mathsf{R}}
\def\B{\textbf {\textup{B}}}
\def\C{\textbf {\textup{C}}}
\def\D{\textbf {\textup{D}}}
\def\X{\textbf {\textup{X}}}
\def\Y{\textbf {\textup{Y}}}
\def\E{\textbf {\textup{E}}}
\def\T{\textbf {\textup{1}}}
\def\A{\textbf {\textup{A}}}
\def\M{\mathcal{M}}
%categorie varie
\newcommand{\catname}[1]{\textbf{\textup{#1}}}
\newcommand{\lab}{\catname{LHyp}}
\newcommand{\lmo}{\catname{LMH}}
\newcommand{\hyp}{\catname{Hyp}}
\newcommand{\phyp}{\catname{pHyp}}
\newcommand{\pshyp}{\catname{pSep}}
\newcommand{\shyp}{\catname{Sep}}
\newcommand{\hyps}{\catname{Hyp}_{\Sigma}}
\newcommand{\EqHyp}{\catname{Hyp}_{\mathsf{eq}}} %equivalence hypergraphs
\newcommand{\EqsHyp}{\catname{Sep}_{\mathsf{eq}}} 
\newcommand{\EqpHyp}{\catname{pHyp}_{\mathsf{eq}}}
\newcommand{\EqpsHyp}{\catname{pSep}_{\mathsf{eq}}}
\newcommand{\Eqlmo}{\catname{LMH}_{\mathsf{eq}}}
\newcommand{\EqHyps}{\catname{Hyp}_{\mathsf{eq}, \mathcal{G}}}
\newcommand{\EqTG}{\catname{EqTG}}
\newcommand{\EqTGs}{\catname{EqTG}_{\Sigma}}
\newcommand{\GEqTGs}{\catname{GEqTG}_{\Sigma}}
\newcommand{\GEqATGs}{\catname{GEqATG}_{\Sigma}}
\newcommand{\gr}{\textbf{\textup{Graph}}}
\newcommand{\dgr}{\catname{SGraph}}
\newcommand{\dg}{\catname{DAG}}
\newcommand{\rt}{\mathsf{dcl_s}}
\newcommand{\rta}{\mathsf{dcl}}
\newcommand{\rtd}{\mathsf{dcl_{d}}}
\newcommand{\slice}[2]{(\catname{#1}\downarrow{#2})}
\newcommand{\tg}[0]{\catname{TG}_{\Sigma}}
\newcommand{\teg}[0]{\catname{TeGr}_{\Sigma}}
\newcommand{\sv}[0]{\mathsf{Sieves}}
\newcommand{\mono}[1]{\mathsf{Mon}(\catname{#1})}
\newcommand{\mo}[1]{{#1}_\mathsf{Mon}}
\newcommand{\pro}{\mathsf{prod}}
\newcommand{\spro}{\mathsf{ps}}
\newcommand{\prol}{\mathsf{lprod}}
\newcommand{\pred}[1]{{\downarrow}#1}
\newcommand{\colim}[0]{\mathrm{colim}}
\newcommand{\cod}{\mathsf{cod}}
\renewcommand{\sp}{\mathsf{sp}}
\newcommand{\pr}{\mathsf{pr}}
\newcommand{\is}[1]{{#1}_{is}}
\renewcommand{\sup}{\mathsf{sup}}
\newcommand{\cow}[1]{\mathsf{cwd}({#1})}
\renewcommand{\inf}{\mathsf{inf}}
\newcommand{\dom}{\mathsf{dom}}
\newcommand{\dwnarrow}{\downarrow \hspace{-2pt}}
\newcommand{\Dwnarrow}{\Downarrow \hspace{-2pt}}
\newcommand{\egg}{\mathsf{e}\text{-}\catname{EqHyp}}

@@

\noindent 
\begin{minipage}[l]{.85\linewidth}
\begin{defi} A \emph{kernel pair} for an arrow $f\colon A \to B$ is an object $K_f$ together with two arrows $\pi^1_f, \pi^2_f\colon K_f \rightrightarrows A$, denoted as $(K_f, \pi^1_f, \pi^2_f)$, such that the square aside is a pullback.
\end{defi}
\end{minipage} \hfill 
\begin{minipage}[r]{.18\linewidth}
	\xymatrix@R=15pt{K_f \ar[r]_{\pi^2_f} \ar[d]_{\pi^1_f}& A \ar[d]^f \\ A \ar[r]^{f}& B}
\end{minipage}

@@

\smallskip 
\noindent
\begin{minipage}[l]{.75\linewidth}
	\begin{defi}\label{def:alg}
		A \emph{(single typed) monoidal signature} $\Sigma$ is a triple $(O_\Sigma, \ari_\Sigma, \coar_\Sigma)$ given by a \emph{set of operations} $O_\Sigma$ together with \emph{arity} and \emph{coarity} functions $\ari_\Sigma, \coar_\Sigma\colon O_\Sigma \rightrightarrows \mathbb{N}$.

@@

\subsubsection{Some subcategories of $\hyp$}\label{sub:sub}

We are now going to introduce some classes of hypergraphs, and the relative subcategories, which will provide the base for our definitions of EGGs  (cfr.~\Cref{sec:sepegg,sec:lmegg}).

@@

\begin{cor}\label{cor:pb}	$\lmo$ is closed in $\hyp$ under pullbacks.
\end{cor}

@@

\paragraph{Pruned and separated hypergraphs}
We can now introduce another  subcategory of $\hyp$ in which we are interested in and that is going to be used in \Cref{sec:sepegg} to give the underlying hypergraphs for EGGs with minimally identified tokens.

@@

 \begin{rem}\label{rem:count}
 	By \Cref{rem:elem2} it is clear that separated hypergraphs enjoy a property similar to the one enjoyed by left-monogamous ones (cfr.~\Cref{proof:count}): if $\mathcal{G}$ is separated and $(f,g)\colon \mathcal{H}\to \mathcal{G}$ is a mono, then $\mathcal{H}$ is separated too.
 \end{rem}

@@

We define the \emph{separated hypergraph associated to $\mathcal{G}$}  as $\sp(\mathcal{G}) = (E_{\mathcal{G}}, V_{\sp(\mathcal{G})}, s_{\sp(\mathcal{G})}, t_{\sp(\mathcal{G})})$.
\end{defi}
\begin{rem}\label{rem:ob}
 By construction $\sp(\mathcal{G})$ is an object of $\shyp$. To see this it is enough to apply \Cref{rem:elem2}
 \begin{itemize}
 	\item every letter of $s_{\mathcal{G}}(e)$ and $t_{\mathcal{G}}(e)$ lies in the image of a different coprojection $\iota_e\circ \iota^e_i$, so it cannot appear more than one time;
 	\item every letter of $s_{\mathcal{G}}(e)$ lies in the image of $\iota_{e}\circ \iota^e_i$  for $i\leq n_e$, while every letter of $t_{\mathcal{G}}(e)$ lies in the image of $\iota_{e}\circ \iota^e_j$  for $n_e < j$, guaranteeing disjointness;
 	\item if $e$ is different from $e'$, then all the letters of $s_{\mathcal{G}}(e)$ and $t_{\mathcal{G}}(e)$ are  in the image of $\iota_e$, while those of  $s_{\mathcal{G}}(e')$ and $t_{\mathcal{G}}(e')$ are in the image of $\iota_{e'}$ and so, again, we get the required disjointness. 
 \end{itemize}
\end{rem}

@@

\begin{rem}\label{rem:mprun}
	Let $(f,g)$ and $(h, k)$ be two morphisms $\mathcal{G}\rightrightarrows \mathcal{H}$ of $\hyp$ with $\mathcal{G}$ pruned and suppose that $f=h$. We can use the absence of isolated nodes in $\mathcal{G}$ to show that $g=k$ too. Let $v\in V_{\mathcal{G}}$, we have two cases:
	\begin{itemize}
		\item $v$ occurs in $s_{\mathcal{G}}(e)$ for some $e\in E_{\mathcal{G}}$. Let $1\leq i\leq \lgh_{V_{\mathcal{G}}}(s_{\mathcal{G}}(e))$ be such that $v=s_{\mathcal{G}}(e)_i$, then we have
		\[g(v)=g(s_{\mathcal{G}}(e)_i)=(g^\star(s_{\mathcal{G}}(e)))_i=s_{\mathcal{H}}(f(e))_i=s_{\mathcal{H}}(h(e))_i=(k^\star(s_{\mathcal{G}}(e)))_i=k(s_{\mathcal{G}}(e)_i)=k(v)\]
		\item $v$ occurs in $t_{\mathcal{H}}(e)$ for some $e\in E_{\mathcal{H}}$. Let $1\leq i\leq \lgh_{V_{\mathcal{H}}}(t_{\mathcal{G}}(e))$ be such that $v=t_{\mathcal{G}}(e)_i$, then we have
		\[g(v)=g(t_{\mathcal{G}}(e)_i)=(g^\star(t_{\mathcal{G}}(e)))_i=t_{\mathcal{H}}(f(e))_i=t_{\mathcal{H}}(h(e))_i=(k^\star(t_{\mathcal{G}}(e)))_i=k(t_{\mathcal{G}}(e)_i)=k(v)\]
	\end{itemize}
	In both cases $\hat{g}(v)=k(v)$ and so $\hat{g}= k$ as wanted.
\end{rem}

\begin{restatable}{prop}{dx}\label{prop:dx}The following hold true:
	\begin{enumerate}
		\item the inclusion functor $P\colon \phyp \to \hyp$ has a right adjoint $\pr$;
		\item The inclusion functor $S_p\colon \pshyp\to \phyp$ has a right adjoint $\sp_P$.
	\end{enumerate}
\end{restatable}

@@

\smallskip
If we restrict to pruned separated hypergraph the previous issue disappeares.  The crucial observation is that the a pruned separated hypergraph is characterized by its set of hyperedges and the length of their sources and targets. See \Cref{proof:pruned} for a proof.

@@

We can use \Cref{lem:pruned} to say something more about colimits and pullbacks in $\pshyp$. A proof can be found in \Cref{proof:limpsy}

\begin{restatable}{prop}{limps}\label{prop:limpsy}
	$\pshyp$ is closed in $\hyp$ under colimits and pullbacks.
\end{restatable}

@@

\begin{defi}
	Let $\mathcal{G}$ be an hypergraph and $(h,k)\colon \mathcal{H}\to \mathcal{G}$  a $\mathcal{G}$-hypergraph. We say that $(h,k)$ is a \emph{left-monogamous} (resp.~\emph{separated, pruned}) $\mathcal{G}$-hypergraph if $\mathcal{H}$ is left-monogamous (resp.~separated, pruned).

	We will denote by $\lmo_{\mathcal{G}}$, $\phyp_{\mathcal{G}}$, $\shyp_{\mathcal{G}}$ and $\pshyp_{\mathcal{G}}$ the corresponding subcategories of $\hyp_{\mathcal{G}}$.
\end{defi}

\begin{rem}Of the four subcategories  defined above, we are interested only in  $\lmo_{\mathcal{G}}$, $\shyp_{\mathcal{G}}$ and $\pshyp_{\mathcal{G}}$. This is due to the fact that we do not have an adhesivity result for $\phyp$.
\end{rem}

\begin{rem} The functors $\pr$ and $\sp_P$ induce right adjoints to the inclusions $P_\mathcal{G}\colon \phyp_\mathcal{G} \to \hyp_{\mathcal{G}}$ and $S_{p, \mathcal{G}}\colon \pshyp_\mathcal{G}\to \phyp_{\mathcal{G}}$.
\end{rem}
We can easily realize $\lmo_{\mathcal{G}}$ and $\shyp_{\mathcal{G}}$ as comma categories.

\begin{prop}
Let $M\colon \lmo \to \hyp$ and $S\colon \shyp \to \hyp$ be the inclusion functors. Let, moreover, $\Gamma_{\mathcal{G}}\colon 1\to \hyp$ be the constant functor in $\mathcal{G}$. Then
\begin{enumerate}
	\item $\lmo_{\mathcal{G}}$ is equivalent to $\comma{M}{\Gamma_\mathcal{G}}$;
	\item $\shyp_{\mathcal{G}}$ is equivalent to $\comma{S}{\Gamma_\mathcal{G}}$; 
	\item $\pshyp_{\mathcal{G}}$ is equivalent to $\pshyp/\sp(\pr(\mathcal{G}))$. 
\end{enumerate}
\end{prop}

@@

\begin{thm} Let $\mathcal{G}$ be an hypergraph, then %the following fact hold
%	\begin{enumerate}
%		\item an arrow $(f,g)$  if and only if is so in $\lmo$; \marginpar{ovvero?}
		%\item 
		$\lmo_{\mathcal{G}}$ is quasiadhesive and $\pshyp_{\mathcal{G}}$ is adhesive.
%	\end{enumerate}
\end{thm}

@@

\noindent 
\begin{minipage}[l]{.5\linewidth}
	\begin{rem}\label{rem:map1}
		Now that we have introduced all the flavours of hypergraphs we need, it is worthwhile to give a visual map of all the categories that we have discussed/introduced and of the relationships between them. We will use the curved arrows to denote full and faithful inclusions, while the vertical arrows are forgetful functors.
	\end{rem}
\end{minipage}\hfill
\begin{minipage}[r]{.45\linewidth}
	\xymatrix@R=8pt@C=10pt{& \pshyp_\mathcal{G} \ar[dd]_(.7){W_\mathcal{G}}|\hole \ar@/_.2cm/@{_{(}->}[rr]_{S_{p,\mathcal{G}}} \ar@{_{(}->}[dl]_{I_\mathcal{G}} && \phyp_\mathcal{G} \ar[dd]^(.7){V_\mathcal{G}}|\hole \ar@/_.2cm/[ll]^{\sp_\mathcal{G}}\ar@/_.2cm/@{_{(}->}[dr]_(.35){P_\mathcal{G}}\\\shyp_\mathcal{G} \ar@{^{(}->}[rr]^(.35){S_\mathcal{G}} \ar[dd]_{R_\mathcal{G}}&& \lmo_\mathcal{G}  \ar@{^{(}->}[rr]^(.35){M_\mathcal{G}} \ar[dd]_(.25){O_\mathcal{G}}&& \hyp_\mathcal{G} \ar@/_.2cm/[ul]_{\pr_\mathcal{G}} \ar[dd]^{U_\mathcal{G}}\\& \pshyp \ar@/_.2cm/@{_{(}->}[rr]_(.35){S_p}|\hole \ar@{^{(}->}[dl]_{I} && \phyp \ar@/_.2cm/[ll]^(.35){\sp}|\hole\ar@/_.2cm/@{_{(}->}[dr]_(.3){P}\\\shyp \ar@{_{(}->}[rr]^{S} && \lmo  \ar@{_{(}->}[rr]^(.35){M}&& \hyp \ar@/_.2cm/[ul]_{\pr}}
\end{minipage}

@@

\Cref{thm:slice-functors,prop:mono}, \Cref{lem:reg} and \Cref{prop:push} allow us to recover the following result, previously proved by direct computation in \cite[Thm.~4.2]{CorradiniG05} (see also \cite[Cor.~5.15]{CastelnovoGM24} for the details of the argument presented here).
\begin{cor}\label{cor:term}
	The category $\tg$ is quasiadhesive.
\end{cor}
\fi 
\section{Adding equivalences to hypergraphical structures}
\label{hypereq}
The use of hypergraphs equipped with an equivalence relation over their nodes has been argued as a convenient way to express concurrency in the DPO approach to rewriting~\cite{concur2006}.
This section presents the framework through adhesive categories, including its extension to the various kinds of hypergraphs introduced in the previous section, as a stepping stone towards an analogous characterization for e-graphs.

@@

\begin{restatable}{lem}{pbsa}\label{lemma:stab}
	In $\EqHyp$, $\pbc$-pushouts are stable.
\end{restatable}

\begin{exa}
It is noteworthy to show that pushouts along regular monos in $\EqHyp$ are not stable.

@@

		\indent As a counterexample, let us consider the cube aside, where the morphisms are the inclusions defined by the labels of the nodes.  The vertices are just graphs without edges, and in the graphs themselves the equivalence classes 

@@

the top face fails to be a pushout.

@@

                \xymatrix@C=10pt@R=.05pt{

@@

\subsubsection{Some subcategories of $\EqHyp$}

We can now focus to the various kind of hypergraphs introduced in \Cref{sub:sub}.

\begin{defi}Given an hypergraph with equivalence $\mathcal{H}:=(E_\mathcal{G}, V_{\mathcal{G}}, Q_\mathcal{G}, s_\mathcal{G}, t_\mathcal{G}, q_\mathcal{G})$, we say that $\mathcal{H}$ is \emph{left-monogamous} (resp.~\emph{separated, pruned}) if $T(\mathcal{H})$ is left-monogamous (resp.~\emph{separated, pruned}). We will denote by $\Eqlmo$, $\EqsHyp$ and $\EqpsHyp$ the  subcategories  given by left-monogamous, separated, pruned and separated hypergraphs with equivalence.
\end{defi}

\begin{rem}\label{rem:obv1}
By definition, there exist functors $a$ as in the diagrams below, where the horizontal arrows are full and faithful inclusions.
\[\xymatrix{\Eqlmo \ar[r]^-{M_{\mathsf{eq}}} \ar[d]_{O_{\mathsf{eq}}}& \EqHyp\ar[d]^{T} & \EqsHyp \ar[r]^-{S_{\mathsf{eq}}} \ar[d]_{R_{\mathsf{eq}}} & \EqHyp \ar[d]^{T} & \EqpsHyp \ar[r]^-{S_{p, \mathsf{eq}}} \ar[d]_{W_{\mathsf{eq}}} & \EqHyp\ar[d]^{T}\\ \lmo \ar[r]_{M} & \hyp & \shyp \ar[r]_{S} & \hyp  & \pshyp \ar[r]_-{S_p}& \hyp}\]

Let us start by dealing with some class of limits in these subcategories (see \Cref{proof:equ} for a proof).

\begin{restatable}{lem}{equ}\label{lem:equ} The following are true:
	\begin{enumerate}
		\item 	$\Eqlmo$ is closed in $\EqHyp$ under equalizers, binary products and pullbacks.
		\item $\EqsHyp$ is closed in $\EqHyp$ under connected limits; 
		\item $\EqpsHyp$ is closed in $\EqHyp$ under pullbacks.
	\end{enumerate}
\end{restatable}

We can now turn to colimits, applying the same strategy. A proof is deferred to \Cref{proof:coequ}

\begin{restatable}{lem}{coequ}\label{lem:coequ}
 The following are true:
\begin{enumerate}
	\item 	$\Eqlmo$ is closed in $\EqHyp$ under pushouts along morphism whose image through $T$ is downward-closed.
	\item $\EqpsHyp$ is closed in $\EqHyp$ under colimits.
\end{enumerate}
\end{restatable}

We are now ready to prove two adhesivity results for $\Eqlmo$ and $\EqpsHyp$.

\begin{defi}We define the following two classes of arrows:
\begin{itemize}
	\item $\pbl$ is the class of all arrows in $\Eqlmo$ sent by $M_{\eq}$ to $\pbc$;
	\item $\pbs$ is the class of all arrows in $\EqpsHyp$ sent by $S_{p,\eq}$ to $\pbc$.
\end{itemize}
\end{defi}

By definition, \Cref{lem:pbmono} and putting together Lemmas \ref{lem:equ}, \ref{lem:coequ} with \Cref{thm:slice-functors} we get the following.

\begin{cor}$\Eqlmo$ is $\pbl$-adhesive and $\EqpsHyp$ is $\pbs$-adhesive
\end{cor}

@@

\begin{defi}
	Let $\mathcal{G}$ be an hypergraph, a \emph{labelled hypergraph with equivalence} is a pair $(\mathcal{H}, (f,g))$ where $\mathcal{H}$ is an object of $\EqHyp$ and $(f,g)$ a morphism $T(\mathcal{H})\to \mathcal{G}$ of $\hyp$. A \emph{morphism of labelled hypergraphs with equivalence} between $(\mathcal{H}, (f,g))$ and $(\mathcal{H}', (h,k))$ is an arrow $(a,b, c)\colon \mathcal{H}\to \mathcal{H}'$ such that:
	\[h\circ a = f \qquad k\circ b = g \].

	We denote the resulting category by $\EqHyps$.
\end{defi}

\noindent 
\begin{minipage}[l]{.8\linewidth}
		\setlength{\parindent}{1.5em}
		\begin{rem}Let $\Sigma$ be a single typed monoidal signature (cfr~\Cref{def:alg}). Since $\mathcal{G}_{\Sigma}$ has a single node, then to label another hypergraph $\mathcal{H}$ it is enough to provide a function $l\colon E_{\mathcal{H}}\to O_{\Sigma}$ such that the diagrams aside commute.

			Notice that, if $\Sigma$ is algebraic, than such a function exists only if $\mathcal{H}$ is algebraic too. In that case the second diagram commute by default.
		\end{rem}
\end{minipage} \hfill 
\begin{minipage}[r]{.15\linewidth}
	\xymatrix{xy code}
	\end{minipage}

Let $(\mathcal{H}, l)$ be an object of $\EqHyps$: since $T$ has a right adjoint $R$ by \Cref{prop:forghyp}, $(f,g)\colon T(\mathcal{H})\to \mathcal{G}^\Sigma$ corresponds to the arrow $(f, g, !_{Q_{\mathcal{H}}})\colon \mathcal{H}\to R(\mathcal{G})$. It is immediate to see that this correspondence extends to an equivalence with the slice category over $R(\mathcal{G})$.

\begin{prop}\label{prop:slice}
$\EqHyps$ is equivalent to $\EqHyp/R(\mathcal{G})$.
\end{prop}

Let $V_\Sigma\colon \EqHyps\to \EqHyp$ be the functor forgetting the labelling and $\pbc_\Sigma$ the class of morphisms in $\EqHyps$ whose image in $\EqHyp$ is in $\pbc$. From \Cref{prop:eqhyp_complete,cor:pbapp}, the second point of \Cref{thm:slice-functors,cor:eqade}, we can deduce the following.
